How has the Brazilian Amazon been constructed as a problem? Presidential speeches and transnational politics since 1985

### 2022-09-10 Draft_September_2022/Full_draft_20220901.rmd
#lastchangesfromworddraft

---
title: How has the Brazilian Amazon been constructed as a problem? Presidential speeches and transnational politics since 1985
author:
- Livio Silva-Muller
- Henrique Sposito^[Both co-authors contributed equally to the article. Names are ordered alphabetically. Livio is a PhD Candidate in Anthropology and Sociology at the Geneva Graduate Institute and affiliated to the Albert Hirschman Centre on Democracy. Henrique is a PhD Candidate in International Relations and Political Sciences at the Geneva Graduate Institute and affiliated to the Centre for International Environmental Studies.]
date: "9 September 2022"
output:
  html_document:
    df_print: paged
  latex_engine: xelatex
  fig_caption: yes
  pdf_document: default
fontsize: 12pt
mainfont: Times New Roman
header-includes:
   - \usepackage{floatrow}
   - \floatsetup[figure]{capposition=top}
   - \usepackage{float}
   - \floatplacement{figure}{H}
   - \usepackage[T1]{fontenc}
   - \usepackage[utf8]{inputenc}
   - |
    \providecommand{\keywords}[1]{\textbf{Keywords:} #1}
    \providecommand{\wordcount}[1]{\text{Word Count:}#1} 
always_allow_html: TRUE
bibliography: references.bib
csl: chicago.csl
abstract: Presidential speeches influence the ways we think about and act towards the environment transnationally, but they are understudied. We propose a framework to investigate how the Brazilian Amazon has been constructed as a problem in discourses along time and space. Using supervised machine learning, we classify statements about the Amazon in 6130 presidential speeches since 1985. We find that national and international events drive the frequency at which the topic of the Amazon is mentioned in presidential speeches inside and outside Brazil. While constructing the Amazon as a problem of economic integration dominates discourses until the mid-2000s, environmental conservation and social development constructions increase in relevance and temporarily surpass economic integration from 2010 to 2015. In turn, constructing the Amazon as an issue of sovereignty steadily increases since 2010. Lastly, the farthest away presidents are from the Amazon itself, the more likely they are to construct the Amazon as an issue of environmental conservation.
---

@@

*We need to protect the Amazon from foreign interests. We need to exploit the Amazon’s natural resources. We need to provide better living standards for the people in the Amazon. We need to preserve the Amazon as a standing ecosystem.* Each of these statements contains an implicit assumption of what needs to be solved, or in other words, it represents the Amazon as a particular problem. These specific problems touch on shared imaginaries of the Amazon as part of the larger socio-cultural history of the country, for example, in relation to issues of national sovereignty, economic integration, social development, or environmental conservation.

Different governments throughout Brazilian history have been described as proponents of a specific view of the Amazon [@drummond2006; @padua2012; @franchini2019; @capobianco2019; @pereira2021]. We conceptualize the Amazon region, forest, and people to be a policy object, this is a specific issue that deserves dedicated policy attention. While the military dictatorship is associated with understanding the Amazon as issues of national sovereignty and economic integration, Sarney’s and Lula’s presidencies are often tied to environmental conservation views. These monolithic representations advance that specific governments understand and act towards the Amazon as one problem, hiding within government diversity of both policies and perspectives. Albeit the current calls to understand the environment as a social-cultural construction and to identify the effect of culture on environmental outcomes [@lepolaindewaroux2021], we lack empirical accounts of how the Brazilian Amazon has been constructed as a problem in discourses over time, by geographical location, and between, or within, governments. Building on @hirschman1963 conceptualization of chosen problems and @bacchi2009 theory of problem-representation in policy, we propose a framework for identifying problem-constructions in discourse and investigate how the Brazilian Amazon has been constructed as a problem in transnational political discourses. We understand transnationalism^[ We use the noun transnationalism, the adverb transnationally, and the adjective transnational interchangeably] as relationships that transcend nation-states, from the local level to the supranational level encompassing non-state actors [@keck1998]

@@

An interesting context to study the relationship between problem-construction and policy objects is presidential speeches, as they legitimize ways of thinking about an issue. While the relationship between presidential speeches and policy may not be causal, they have the power to introduce and justify public policy, as well as shape issue perception to broad audiences [@zarefsky2004; @gillion2016]. As presidents speak across several locations to various audiences, they frame similar issues differently.

The possibility of different problem-constructions in presidential speeches has important implications for democracy. With globalization and the increase of transnational processes, presidents have the opportunity of promoting different problem-constructions transnationally. It is important to understand, thus, to what extent the implemented policy agenda correlates with problem-constructions promoted at the local, federal, or international levels. Relatedly, when Brazilian presidents speak about the Amazon it not only makes headlines, nationally and internationally [@brice2021; @harris2021; @miranda2021], but also incites responses, shapes expectations, and feeds into the behavior of a myriad of related actors. Deforestation rates in Brazil are more responsive to the government’s environmental policy, generated from material and discursive governmental practices than exogenous factors such as market fluctuations [@assuncao2015; @capobianco2019; @capobianco2021]. Specifically, @capobianco2021 finds that the perception of a higher risk of being caught for deforestation in the Amazon is the main mechanism through which environmental policy is effective. Furthermore, ethnographic research has also shown that expectations about governments’ responses are a crucial factor in decisions to deforest on the ground [@campbell2015]. Understanding variation in presidential speeches is the first step to connecting it to concrete environmental outcomes.

@@

This variation in problem-construction entails that the same governments are more diverse in their positions than the literature suggests. Scholars often describe federal governments as proponents of a cohesive set of policies toward the Amazon [see @drummond2006; @padua2012; @franchini2019; @capobianco2019; @pereira2021]. In this literature, the 1964 military dictatorship is associated with securing sovereignty in the region by populating it and integrating it into the national economy [@drummond2006]. The governments from the late 1980s up to 2009 are associated with a turn towards policies that focus on environmental conservation. The presidencies of Dilma and Temer are connected to the decline of conservationist policies, while Bolsonaro to the dismantling of environmental policies. Although these works are important to understand how different governments have acted toward the Amazon, the classification of specific governments into policy periods, or policy cycles, represents them as monolithic: they associate specific governments with one specific view of the Amazon.

@@

Second, with the codebook in hand, each one of the authors, separately, hand-coded the same set of 1007 randomly selected Amazonian statements. This amount refers to 50% of all the Amazon Statements identified. We chose to hand.code half of the observations because since there are several nuances in discourse in how presidents talk about the Amazon over time and as a problem, we conservatively coded a relatively large training set for the subsequent automated machine learning. As well, this allows a robust validation set to verify the models. As the size of the training set should increase with the number of categories [@grimmer2022], we deemed four categories and 50% of the statements as reliable. Also, automating the coding of half of the observations saved the authors over one month of work in comparison to manual coding. The intercoder agreement for each of the four main categories was 85%, on average. For each non-matching coded observation, the co-authors discussed and sorted their disagreements.

@@

We observe five local maxima both inside and outside Brazil: 1989, 1992, 2009, 2015, and 2019. These points coincide with internal and external events that help us explain the rises and falls of the Amazon in presidential discourse. From 1985 to 1988, the mentions of the Amazon in official speeches increased from 4% to 14%. During this period, the latest Brazilian Constitution, which gave indigenous and traditional populations space to advocate for constitutional environmental rights and the protection of their territories [@hecht1990], was written. In 1989, however, the Amazon appeared in 32% of all speeches. This spike correlates with the brutal murder of Chico Mendes in the last days of 1988. The incident caught unprecedented international attention and Sarney responded to this with a set of policies to address deforestation [@capobianco2021], including accepting to host the 1992 Earth Summit [@keck1998].

1990-91 the Amazon appeared in 12% to 14% of all speeches but, by 1992, this average increased to 25%. The 1992 Earth Summit in Brazil brought international attention to environmental topics related to the Amazon. One of the big announcements, for example, was the consolidation of the first transnational partnership for the Amazon, the G7 Pilot Program, which brought a high number of financial resources to the region for public policy implementation [@capobianco2021]. Throughout the rest of the 1990s and 2000s, during the Cardoso and Lula administrations, the Amazon as a topic did not diverge much from the average until 2009, when we see another spike in mentions of the Amazon. This coincides with the 2009 Copenhagen Summit and the steepest decrease in deforestation rates since data is available as Figure 3, below, illustrates. Lula led the delegation to Copenhagen with a self-image of “we do not promise, we deliver” when the stakes about climate change were high [@franchini2019].

@@

From 2010 to 2018, except for 2015, we see a general decrease in mentions of the Amazon in all official presidential speeches. From 2010 to 2014, mentions of the Amazon went down from 18% to 8%. Disagreements related to the priority of environmental preservation over economic development, most notably in the case of approval of the New Forest Code in 2012, which regularized land ownership of many illegally deforested areas, were frequent. The period was also marked by political and economic instability which culminated in the impeachment of Rousseff in 2016. In 2015, the Amazon appeared in 19% of all speeches. This spike coincides with the Paris summit which became a key turn in climate politics after the failures of Copenhagen. Brazil went to the Paris Summit with deforestation numbers slightly higher than Copenhagen and a perception that there was a turn towards less conservation inside the country. By 2016 mentions of Amazon in official speeches went down to 6%, the lowest share since 1985.

@@

We conceptualize four problem-constructions: sovereignty, economic integration, social development, and conservation. Figure 4, below, illustrates the share of each pure-type in time. Pure problem-constructions dominate, with an average of 56% of all Amazonian statements across time.We identify a discursive shift in problem-constructions from developing the Amazon economically, to conserving it and providing services to its citizens. Pure economic integration constructions dominated constructions from 1985 to the late 2000s. This is especially pertinent during the Cardoso administration. Brazilian presidents after Cardoso moved from employing one problem-constructions consistently to a more diverse strategy. During Lula’s terms in office, constructing the Amazon purely as an issue of economic integration decreased, while constructing the Amazon issues of environmental conservation and social development increased. Conservation and social development constructions surpassed economic integration around 2010 and continued to dominate as Amazonian constructions during Rousseff's administration. Still, environmental conservation and social development problem-constructions were not as dominant in terms of frequency or time in discourses as economic integration was until the early 2000s. Finally, we observe a steady increase in sovereignty constructions in the late 2000s, reaching the highest frequencies with the presidencies of Bolsonaro.

@@

Our models also suggest that deforestation plays an important role. Presidents are, for example, more likely to construct the Amazon as an issue of environmental conservation as deforestation decreases. Alternatively, as deforestation increases, presidents are more likely to construct the amazon as an issue of economic integration. This indicates that Brazil creates a self-image of a strong climate leadership because of the good results achieved from 2004 to 2012 [@franchini2019]. In the same fashion, economic integration problem-constructions are likely deployed as a justification for increasing deforestation rates.

@@

The steady increase of sovereignty problem-constructions starting in 2010 brings an eerie twist. As we conceptualize and operationalize sovereignty as containing internal and external perceived threats to the Amazon, we interpret this increase as not only attacks on international interference in the Amazon but also on indigenous and traditional populations. The embryo of Bolsonaro’s Amazonian discourse was breeding half a decade before he took office. Bolsonaro is a product of political re-organization of national and international actors - to which Rousseff and Temer were already responding in discourse - rather than the cause of change in politics. The political forces in Brazilian democracy that drove these policy and discursive changes in problem-construction were long in the making, Bolsonaro’s problem-constructions and policies are the strongest forms of this shift.

>>>>

---
title: How has the Brazilian Amazon been constructed as a problem? Presidential speeches and transnational politics since 1985
author:
- Livio Silva-Muller
- Henrique Sposito^[Both co-authors contributed equally to the article. Names are ordered alphabetically. Livio is a PhD Candidate in Anthropology and Sociology at the Geneva Graduate Institute and affiliated to the Albert Hirschman Centre on Democracy. Henrique is a PhD Candidate in International Relations and Political Sciences at the Geneva Graduate Institute and affiliated to the Centre for International Environmental Studies.]
date: "9 September 2022"
output:
  pdf_document: default
  latex_engine: xelatex
  fig_caption: yes
  html_document:
    df_print: paged
fontsize: 12pt
mainfont: Times New Roman
header-includes:
   - \usepackage{floatrow}
   - \floatsetup[figure]{capposition=top}
   - \usepackage{float}
   - \floatplacement{figure}{H}
   - \usepackage[T1]{fontenc}
   - \usepackage[utf8]{inputenc}
   - |
    \providecommand{\keywords}[1]{\textbf{Keywords:} #1}
    \providecommand{\wordcount}[1]{\text{Word Count:}#1} 
always_allow_html: TRUE
bibliography: references.bib
csl: chicago.csl
abstract: Presidential speeches influence the ways we think about and act towards the environment transnationally, but they are understudied. We propose a framework to investigate how the Brazilian Amazon has been constructed as a problem in discourses across time and space. Using supervised machine learning, we classify statements about the Amazon in 6130 presidential speeches since 1985. We find that national and international events drive the frequency at which the topic of the Amazon is mentioned in presidential speeches inside and outside Brazil. While constructing the Amazon as a problem of economic integration dominates discourses until the mid-2000s, environmental conservation and social development constructions increase in relevance and temporarily surpass economic integration from 2010 to 2015. In turn, constructing the Amazon as an issue of sovereignty steadily increases since 2010. Lastly, the farthest away presidents are from the Amazon itself, the more likely they are to construct the Amazon as an issue of environmental conservation.
---

@@

*We need to protect the Amazon from foreign interests. We need to exploit the Amazon’s natural resources. We need to provide better living standards for the people in the Amazon. We need to preserve the Amazon as a standing ecosystem.* Each of these statements contains an implicit assumption of what needs to be solved, or in other words, it represents the Amazon as a particular problem. These specific problems touch on common understandings of the Amazon as part of the larger socio-cultural history of the country, for example, in relation to issues of national sovereignty, economic integration, social development, or environmental conservation.

Different governments throughout Brazilian history have been described as proponents of a specific view of the Amazon [@drummond2006; @padua2012; @franchini2019; @capobianco2019; @pereira2021]. We conceptualize the Amazon region, forest, and people to be a policy object, this is a specific issue that deserves dedicated policy attention. While the military dictatorship(1964-85) is associated with understanding the Amazon as issues of national sovereignty and economic integration, Sarney’s  (1985-1990) and Lula’s (2003-2010)presidencies are often tied to environmental conservation views. These monolithic representations advance that specific governments understand and act towards the Amazon as one problem, hiding within government diversity of both policies and perspectives. Albeit the current calls to understand the environment as a social-cultural construction and to identify the effect of culture on environmental outcomes [@lepolaindewaroux2021], we lack empirical accounts of how the Brazilian Amazon has been constructed as a problem in discourses over time, by geographical location, and between, or within, governments. Building on @hirschman1963 conceptualization of chosen problems and @bacchi2009 theory of problem-representation in policy, we propose a framework for identifying problem-constructions in discourse and investigate how the Brazilian Amazon has been constructed as a problem in transnational political discourses. We understand transnationalism^[ We use the noun transnationalism, the adverb transnationally, and the adjective transnational interchangeably] as relationships that transcend nation-states, from the local level to the supranational level encompassing non-state actors [@keck1998].

@@

An interesting context to study the relationship between problem-construction and policy objects is presidential speeches, as they legitimize ways of thinking about an issue. While the relationship between presidential speeches and policy may not be causal, they have the power to introduce and justify public policy, as well as shape issue perception to broad audiences [@zarefsky2004; @gillion2016]. As presidents speak across several locations to various audiences, they construct similar issues differently.

The possibility of different problem-constructions in presidential speeches has important implications for democracy. With globalization, presidents have the opportunity of promoting different problem-constructions transnationally. It is important to understand, thus, to what extent the implemented policy agenda correlates with problem-constructions promoted at the local, federal, or international levels. Relatedly, when Brazilian presidents speak about the Amazon it not only makes headlines, nationally and internationally [@brice2021; @harris2021; @miranda2021], but also incites responses, shapes expectations, and feeds into the behavior of a myriad of related actors. Deforestation rates in Brazil are more responsive to the government’s environmental policy, generated from material and discursive governmental practices than exogenous factors such as market fluctuations [@assuncao2015; @capobianco2019; @capobianco2021]. Specifically, @capobianco2021 finds that the perception of a higher risk of being caught for deforestation in the Amazon is the main mechanism through which environmental policy is effective. Furthermore, ethnographic research has also shown that expectations about governments’ responses are a crucial factor in decisions to deforest on the ground [@campbell2015]. Understanding variation in presidential speeches is the first step to connecting it to concrete environmental outcomes.

@@

This variation in problem-construction entails that the same governments are more diverse in their positions than the literature suggests. Scholars often describe federal governments as proponents of a cohesive set of policies toward the Amazon [see @drummond2006; @padua2012; @franchini2019; @capobianco2019; @pereira2021]. In this literature, the 1964 military dictatorship is associated with securing sovereignty in the region by populating it and integrating it into the national economy [@drummond2006]. The governments from the late 1980s up to 2009 are associated with a turn towards policies that focus on environmental conservation. The presidencies of Dilma and Temer (2011-2018) are connected to the decline of conservationist policies, while Bolsonaro to the dismantling of environmental policies. Although these works are important to understand how different governments have acted toward the Amazon, the classification of specific governments into policy periods, or policy cycles, represents them as monolithic: they associate specific governments with one specific view of the Amazon.

@@

Second, with the codebook in hand, each one of the authors, separately, hand-coded the same set of 1007 randomly selected Amazonian statements. This amount refers to 50% of all the Amazon Statements identified. We chose to hand.code half of the observations because since there are several nuances in discourse in how presidents talk about the Amazon over time and as a problem, we conservatively coded a relatively large training set for the subsequent automated machine learning. As well, this allows a robust validation set to verify the models. As the size of the training set should increase with the number of categories [@grimmer2022], we deemed four categories and 50% of the statements as reliable. Also, automating the coding of half of the observations saved the authors over one month of work in comparison to manual coding. The intercoder agreement for hand-coded categories was 85%, on average. For each non-matching coded observation, the co-authors discussed and sorted their disagreements.

@@

We observe five local maxima both inside and outside Brazil: 1989, 1992, 2009, 2015, and 2019. These points coincide with internal and external events that help us explain the rises and falls of the Amazon in presidential discourse. From 1985 to 1988, the mentions of the Amazon in official speeches increased from 4% to 14%. During this period, the latest Brazilian Constitution, which gave indigenous and traditional populations space to advocate for constitutional environmental rights and the protection of their territories [@hecht1990], was written. In 1989, however, the Amazon appeared in 32% of all speeches. This maximum correlates with the brutal murder of Chico Mendes in the last days of 1988. The incident caught unprecedented international attention and Sarney responded to this with a set of policies to address deforestation [@capobianco2021], including accepting to host the 1992 Earth Summit [@keck1998].

1990-91 the Amazon appeared in 12% to 14% of all speeches but, by 1992, this average increased to 25%. The 1992 Earth Summit in Brazil brought international attention to environmental topics related to the Amazon. One of the big announcements, for example, was the consolidation of the first transnational partnership for the Amazon, the G7 Pilot Program, which brought a high number of financial resources to the region for public policy implementation [@capobianco2021]. Throughout the rest of the 1990s and 2000s, during the Cardoso (1995-2002) and Lula administrations, the Amazon as a topic did not diverge much from the average until 2009, when we see another maximum in mentions of the Amazon. This coincides with the 2009 Copenhagen Summit and the steepest decrease in deforestation rates since data is available as Figure 3, below, illustrates. Lula led the delegation to Copenhagen with a self-image of “we do not promise, we deliver” when the stakes about climate change were high [@franchini2019].

@@

From 2010 to 2018, except for 2015, we see a general decrease in mentions of the Amazon in all official presidential speeches. From 2010 to 2014, mentions of the Amazon went down from 18% to 8%. Disagreements related to the priority of environmental preservation over economic development, most notably in the case of approval of the New Forest Code in 2012, which regularized land ownership of many illegally deforested areas, were frequent. The period was also marked by political and economic instability which culminated in the impeachment of Rousseff in 2016. In 2015, the Amazon appeared in 19% of all speeches. This maximum coincides with the Paris summit which became a key turn in climate politics after the failures of Copenhagen. Brazil went to the Paris Summit with deforestation numbers slightly higher than Copenhagen and a perception that there was a turn towards less conservation inside the country. By 2016 mentions of Amazon in official speeches went down to 6%, the lowest share since 1985.

@@

We conceptualize four problem-constructions: sovereignty, economic integration, social development, and conservation. Figure 4, below, illustrates the share of each pure-type in time. Pure problem-constructions dominate, with an average of 56% of all Amazonian statements across time.We identify a discursive shift in problem-constructions from developing the Amazon economically, to conserving it and providing services to its citizens. Pure economic integration constructions dominated constructions from 1985 to the late 2000s. This is especially pertinent during the Cardoso administration. Brazilian presidents after Cardoso moved from employing one problem-constructions consistently to a more diverse strategy. During Lula’s terms in office, constructing the Amazon purely as an issue of economic integration decreased, while constructing the Amazon issues of environmental conservation and social development increased. Conservation and social development constructions surpassed economic integration around 2010 and continued to dominate as Amazonian constructions during Rousseff's administration. Still, environmental conservation and social development problem-constructions were not as dominant in terms of frequency or periods in time in discourses as economic integration was until the early 2000s. Finally, we observe a steady increase in sovereignty constructions in the late 2000s, reaching the highest frequencies with the presidencies of Bolsonaro.As we conceptualize and operationalize sovereignty as containing internal and external perceived threats to the Amazon, we interpret this increase as not only attacks on international interference in the Amazon but also on indigenous and traditional populations.

@@

Our models also suggest that deforestation plays an important role. Presidents are, for example, more likely to construct the Amazon as an issue of environmental conservation as deforestation decreases. Alternatively, as deforestation increases, presidents are more likely to construct the amazon as an issue of economic integration. This indicates that Brazil created a self-image of a strong climate leadership because of the good results achieved from 2004 to 2012 [@franchini2019]. In the same fashion, economic integration problem-constructions are likely deployed as a justification for increasing deforestation rates.

@@

The steady increase of sovereignty problem-constructions starting in 2010 brings an eerie twist.  The embryo of Bolsonaro’s Amazonian discourse was breeding half a decade before he took office. Bolsonaro is a product of political re-organization of national and international actors - to which Rousseff and Temer were already responding in discourse - rather than the cause of change in politics. The political forces in Brazilian democracy that drove these policy and discursive changes in problem-construction were long in the making, Bolsonaro’s problem-constructions and policies are the strongest forms of this shift.
